CSD, Disjoint components and trayectory detection

### 2019-07-18 ArticuloCSDyCM/Paper03.tex
mas figuras

Center of Mass Analysis (CMA) has been used previously in order to track putative  ``trajectories'' of the activity or information across neural tissue or even the whole brain \cite{Chao05, Chao07, Manjarrez07, Manjarrez09}. In such works the chosen ``mass'' was an heuristic measure adequate for representing the coarse-grain picture of displacement of activity. These present the problem of justifying the choice made on practical grounds. For calculating a vectorial average, the adequate weight per vector is a density, a positive quantity that measures the concentration of something per space unit, so it is only adequate that we search for a natural density in the electrophysiological experiments. The Current Source Density fits the mathematical and interpretative requirements. We shall denote it in the following as $C$ (following \cite{Potworowski2011}).  This measure is usually presented in arbitrary units, which permits us to interpret it two-fold. One, directly from its definition, as the \emph{differential divergence} of the vectorial electric current density:
\begin{equation}
  C:=\nabla \cdot J
\end{equation}
Due to conservation of charges, this is the same as as the temporal variation of density of charge:
\begin{equation}
  C=-\frac{\partial \rho}{\partial t}
\end{equation}
Current Source Density as a mathematical density (a quasi-probability density function)  allows  to rigorously apply the concepts of vector averaging. This use has been overlooked. This might have occurred because  samplings at lesser spatial resolution did not permit estimations of this quantity with enough precision to make reliable spatial implications, even with the use of sophisticated CSD estimation methods. In our case, we want to use the density as a mathematician would: a function which assigns weights to points in the space of interest and permits us to carry linear operations over them.  The space here is the sampling points of the electrodes at a given instant in time, and the weight, the CSD, is amount of interchange of charged ions from the inside to the outside of cellular membranes, per unit space.  This is an indicator of activity in the neurons, therefore the CSD is also, indirectly, a \emph{``density of local activity''}. Having both
the mathematician and electrophysiologicist interpretations presents we can
perform the following analysis with clear readings.

@@

To clarify the concept and how we are we using it an example is in order. In figure \ref{disconnectedsets} A we show geometrical sets enclosed by simple curves as color patches, where every color indicates a Set. The set $A$ consists of a single connected component, that means that one can make  path between any two points in the set that consist entirely of points of the same set. In contrast, the set $B$ is made of two disconnected components, labeled $B1$ and $B2$. If one chooses a point in each of the components, one cannot draw a path joining them without leaving the set. But each one of those components is connected, meaning that if we choose the pair of points in one component, then again we can trace a path joining them in the set. So the set $B$ consists of two disconnected components, but each component can be regarded as a connected set on their own. Likewise the set $C$ is made of three disconnected components. In the subfigure \ref{disconnectedsets} Aiii we show the center of mass of the sets. The set $A$ is concave, so, it can have a center of mass outside the set. The
set $B$ has also a center of mass outside the set. It has two disconnected components that are aproximatelly
equal and separated by a long distance (this is not an implication: a set can be disconnected and have
a CM inside the set). For the set $C$ our approach has been used: we treat every disconnected component of the
set as an independent entity and calculate the center of mass for each one. Notice that this doesn't preclude t
he possibility of having a concave subset with a CM that lies outside. But, as we shall see,
as the electrode resolution is finite, the possibility of having highly non compact shapes is unlikely, and they
tend to have a CM inside or near their borders. In the figure \ref{disconnectedsets} B we show this with experimental
data (we are using evoked activity data, as explained in section \ref{sec:evocada}). In graph Bi we
separate the CSD data into the sink and sources set. We take sinks as example. In the graph Bii
we label each disjoint component, obtained by a single pass algorithm. There are 25 components, and
the bigger two are non convex. In the
subfigure Biii we indicate the CM of only those components that are above certain threshold of
integrated intensity (the circle size  is proportional in area to such intensity, there are five CM to be
seen). Each CM gets listed as a probable step in a trajectory.

@@

@@

Once that we have the CSD representation, we select channels outside
the r.o.i. and use them to obtain the threshold for separating
the Sink and Sources sets. Then we procede as stated in the previous section.

The figure \ref{lfpcsdcm}  compares the representation of the activity in
the LFP, CSD and CM spaces. It has been exhaustively proven that the CSD
representation provides better spatial localization of active sites, but the
CM can serve as better dynamical tracker of action. The figure illustrates
that point by choosing the time interval of response in CA3 after the stimulus.
CM representation permits us to track the activity with less
interference from the saturated region. The small, clearly localized centers of
mass separate clearly from the big patch over the saturated region.
In the figure \ref{evotracks} we concatenate the successive centers of mass
for both sinks and sources and show how the anterograde reaction is
very well localized over the Pyramidal Stratum at precise moments after
stimulus.

The treatment of the data is shown in figure
\ref{CSD_Evocada01}.  The subfigure A  shows a diagram of the hippocampus, with
the somatic layers (CA and DG) as thick black lines and the stimulus point
shown as a blue arrow. The $a,b,c,d,e$ dots are selected electrodes for
showing LFP and CSD traces. Those are shown in the subfigure B. In subfigure
C the same data is presented as a spatiotemporal color map that
shows the expected poles of firing neurons transverse to the
pyramidal stratum.  In the subfigure D we show the CSD for a series of time frames
over the square delimited in the subfigure A. The idea for treating the different
patches separately becomes evident here. The sources and sinks form patches
with an apparent anterograde movement from CA3 to CA1.
The patches change polarity
as expected, with the somatic layer becoming a strong sink between 3 to 4 ms
after the stimulus, and then, at the 6 ms, inverting its polarity in a way that suggest
successive activation of groups that are represented as movement of the patch.

\begin{figure}[h]
  \includegraphics[width=0.9\textwidth]{Figuras/CSD_EvocadaYTrazos.pdf}
  \caption{ CSDA for evoked activity and visual inspection of disjoint
    components. In A we present a diagram of the hippocampus slice over
    the acquisition device. The somatic layers of CA and DG are represented as
    thick black lines, and the stimulus site is indicated by a blue arrow. A selection
    of six contiguous electrodes labeled $A-F$ in the r.o.i. are presented as blue
    dots. In B we show the LFP traces and their CSD representation for those
    selected electrodes.
    In C we shoe the CSDA for various timestamps in the sub-region
    indicated in A as the r.o.i, which covers a square of 24 by 24 electrodes over CA3.
    in the first image,  4ms after the stimulus,  a
    shaded strip indicating the approximate location of str. pyr. A model neuron is
    drawn for scale. At 4ms after the stimulus, the saturation of the electric shock
    recedes and we can began to appreciate real neuronal activity on the responsive
    sector of CA3. A great sink appears over the somatic layer, surrounded by two
    sources and some smaller patches of activity. At the following frame the patch
    appears to have moved anterogradly, towards CA1. Next frame, the sink appears
    to have inverted its polarity as do their surrounding patches, and the movement
    continues until approximately 9ms after stimulus, where the activity ceases to be
    discernible from noise.}
  \label{CSD_Evocada01}
\end{figure}

 \includegraphics[width=0.9\textwidth]{Figuras/CMVariousTimes01.pdf}
+  \caption{ The CM analysis over the same images as figure \ref{CSD_Evocada01} C. Some trajectories
+    are calculated before the 4ms time stamp. The triangular mark shows the later point of the
+    trajectory, either the end or the last calculated position until that moment. The circular
+    mark is the start of the trajectory. We can see that activity moves form CA3 distal to
+    proximal in alternating waves of action (hiperpolarization, firing and repolarization).
+  }\label{cmevocada}

@@

\begin{figure}
  \includegraphics[width=0.9\textwidth]{./Figuras/LFPvsCSD_4AP.pdf}
  \caption{Comparison of LFP and CSD representations
    at the moments of the epileptic burst. In the first row
    we present the data as measured, no denoizing or filtering has
    been applied. The units of measurement are in $\muV{}$. In
    the middle row a direct application of a difference operator
    gives us the CSD. This sort of operators are very sensitive
    to noise, in this case, spatial hard edges, which become
    greatly exaggerated. In the third row we present the same
    data after a Gaussian spatial filter ( Gaussian Blur) has
    been applied. }\label{lfpycsd4ap}
\end{figure}

>>>>

Center of Mass Analysis (CMA) has been used previously in order to track putative  ``trajectories'' of the activity or information across neural tissue or even the whole brain \cite{Chao05, Chao07, Manjarrez07, Manjarrez09}. In such works the chosen ``mass'' was an heuristic measure adequate for representing the coarse-grain picture of displacement of activity. These present the problem of justifying the choice made on practical grounds. For calculating a vectorial average, the adequate weight per vector is a density, a positive quantity that measures the concentration of something per space unit, so it is only adequate that we search for a natural density in the electrophysiological experiments. The Current Source Density fits the mathematical and interpretative requirements. We shall denote it in the following as $C$ (following \cite{Potworowski2011}).  This measure is usually presented in arbitrary units, which permits us to interpret it two-fold. One, directly from its definition, as the \emph{differential divergence} of the vectorial electric current density:
\begin{equation}
  C:=\nabla \cdot J
\end{equation}
Due to conservation of charges, this is the same as as the temporal variation of density of charge:
\begin{equation}
  C=-\frac{\partial \rho}{\partial t}
\end{equation}
Current Source Density as a mathematical density (a quasi-probability density function)  allows  to rigorously apply the concepts of vector averaging. This use has been overlooked. This might have occurred because  samplings at lesser spatial resolution did not permit estimations of this quantity with enough precision to make reliable spatial implications, even with the use of sophisticated CSD estimation methods. In our case, we want to use the density as a mathematician would: a function which assigns weights to points in the space of interest and permits us to carry linear operations over them.  The space here is the sampling points of the electrodes at a given instant in time, and the weight, the CSD, is amount of interchange of charged ions from the inside to the outside of cellular membranes, per unit space.  This is an indicator of activity in the neurons, therefore the CSD is also, indirectly, a \emph{``density of local activity''}. Having both
the mathematician and electrophysiologist interpretations presents we can
perform the following analysis with clear readings.

@@

To clarify the concept and how we are we using it an example is in order. In figure \ref{disconnectedsets} A we show geometrical sets enclosed by simple curves as color patches, where every color indicates a Set. The set $A$ consists of a single connected component, that means that one can make  path between any two points in the set that consist entirely of points of the same set. In contrast, the set $B$ is made of two disconnected components, labeled $B1$ and $B2$. If one chooses a point in each of the components, one cannot draw a path joining them without leaving the set. But each one of those components is connected, meaning that if we choose the pair of points in one component, then again we can trace a path joining them in the set. So the set $B$ consists of two disconnected components, but each component can be regarded as a connected set on their own. Likewise the set $C$ is made of three disconnected components. In the subfigure \ref{disconnectedsets} Aiii we show the center of mass of the sets. The set $A$ is concave, so, it can have a center of mass outside the set. The
set $B$ has also a center of mass outside the set. It has two disconnected components that are approximately
equal and separated by a long distance (this is not an implication: a set can be disconnected and have
a CM inside the set). For the set $C$ our approach has been used: we treat every disconnected component of the
set as an independent entity and calculate the center of mass for each one. Notice that this doesn't preclude t
he possibility of having a concave subset with a CM that lies outside. But, as we shall see,
as the electrode resolution is finite, the possibility of having highly non compact shapes is unlikely, and they
tend to have a CM inside or near their borders. In the figure \ref{disconnectedsets} B we show this with experimental
data (we are using evoked activity data, as explained in section \ref{sec:evocada}). In graph Bi we
separate the CSD data into the sink and sources set. We take sinks as example. In the graph Bii
we label each disjoint component, obtained by a single pass algorithm. There are 25 components, and
the bigger two are non convex. In the
subfigure Biii we indicate the CM of only those components that are above certain threshold of
integrated intensity (the circle size  is proportional in area to such intensity, there are five CM to be
seen). Each CM gets listed as a probable step in a trajectory.

@@

%The treatment of the data is shown in figure
%\ref{CSD_Evocada01}.  The subfigure A  shows a diagram of the hippocampus, with
%the somatic layers (CA and DG) as thick black lines and the stimulus point
%shown as a blue arrow. The $a,b,c,d,e$ dots are selected electrodes for
%showing LFP and CSD traces. Those are shown in the subfigure B. In subfigure
%C the same data is presented as a spatiotemporal color map that
%shows the expected poles of firing neurons transverse to the
%pyramidal stratum.  In the subfigure D we show the CSD for a series of time frames
%over the square delimited in the subfigure A. The idea for treating the different
%patches separately becomes evident here. The sources and sinks form patches
%with an apparent anterograde movement from CA3 to CA1.
%The patches change polarity
%as expected, with the somatic layer becoming a strong sink between 3 to 4 ms
%after the stimulus, and then, at the 6 ms, inverting its polarity in a way that suggest
%successive activation of groups that are represented as movement of the patch.

@@

Once that we have the CSD representation, we select channels outside
the r.o.i. and use them to obtain the threshold for separating
the Sink and Sources sets. Then we proceed as stated in the previous section.

The figure \ref{lfpcsdcm}  compares the representation of the activity in
the LFP, CSD and CM spaces. The CSD
representation provides better spatial localization of active sites, but the
CM can serve as better dynamical tracker of action. The figure illustrates
that point by choosing the time interval of response in CA3 after the stimulus.
CM representation permits us to track the activity with less
interference from the saturated region. The small, clearly localized centers of
mass separate clearly from the big patch over the saturated region.
In the figure \ref{lfpcsdcmsub} we repeat the procedure but focusing in
the r.o.i., which includes CA3 str. pyr. We take advantage of the
detailed frame to show the borders of the disconnected sets that
we have chosen for the CM analysis in the right column.

In the figure \ref{evotracks} we concatenate the successive centers of mass
for both sinks and sources and show how the anterograde reaction is
very well localized over the Pyramidal Stratum at precise moments after
stimulus. This dataset provides us with a very clean example of the
trajectory analysis. The very well localized activity and coherent
action gives us trajectories that are in accordance with the canonical
source and sink distribution, but adding the temporal dynamical displacement
to it. We have been very thorough with this example as it will help us
to understand the more complicated data that shall be presented in
the next examples.

\begin{figure}
  \includegraphics[width=0.9\textwidth]{Figuras/lfpevocado01.pdf}
  \caption{A set of frames showing the stimulus and posterior activity.
    The frame titled ``0 ms'' is just before the stimulus was
    made.  A diagram of the Granular and Pyramidal Layers is shown
    against the noice background. The black rectangle delimites the
    r.o.i. and the black arrow shows the site of stimulation in GD.
  }\label{lfpevocada}
\end{figure}

\begin{figure}[h]
\includegraphics[width=0.9\textwidth]{Figuras/lfpcsdcm01.pdf}
\caption{ Comparison of the activity represented as LFP, CSD and CM.
  The time after
  stimulation in ms is shown in the central column for each row.
  In this series of frames a relatively big patch of the electrodes
  still appear saturated, but even so the CSD shows good localization
  of the relevant deflections from the neutral or noise base signal.
  The CM representation concentrates the locus of action even more.
  The circular dots in the third column represent the center of mass
  of coherent active subsets.
}\label{lfpcsdcm}
\end{figure}

\begin{figure}[h]
\includegraphics[width=0.9\textwidth]{Figuras/lfpcsdcm02.pdf}
\caption{ The same as previous figure, but focusing on the
  region of interest, including the stratum pyramidale of CA3.
}\label{lfpcsdcmsub}
\end{figure}

%\begin{figure}[h]
%  \includegraphics[width=0.9\textwidth]{Figuras/CSD_EvocadaYTrazos.pdf}
%  \caption{ CSDA for evoked activity and visual inspection of disjoint
%    components. In A we present a diagram of the hippocampus slice over
%    the acquisition device. The somatic layers of CA and DG are represented as
%    thick black lines, and the stimulus site is indicated by a blue arrow. A selection
%    of six contiguous electrodes labeled $A-F$ in the r.o.i. are presented as blue
%    dots. In B we show the LFP traces and their CSD representation for those
%    selected electrodes.
%    In C we shoe the CSDA for various timestamps in the sub-region
%    indicated in A as the r.o.i, which covers a square of 24 by 24 electrodes over CA3.
%    in the first image,  4ms after the stimulus,  a
%    shaded strip indicating the approximate location of str. pyr. A model neuron is
%    drawn for scale. At 4ms after the stimulus, the saturation of the electric shock
%    recedes and we can began to appreciate real neuronal activity on the responsive
%    sector of CA3. A great sink appears over the somatic layer, surrounded by two
%    sources and some smaller patches of activity. At the following frame the patch
%    appears to have moved anterogradly, towards CA1. Next frame, the sink appears
%    to have inverted its polarity as do their surrounding patches, and the movement
%    continues until approximately 9ms after stimulus, where the activity ceases to be
%    discernible from noise.}
%  \label{CSD_Evocada01}
% \end{figure}

\begin{figure}
 \includegraphics[width=0.9\textwidth]{Figuras/CMVariousTimes01.pdf}
  \caption{ The CM analysis over the same images as figure \ref{lfpcsdcmsub} C. Some trajectories
    are calculated before the 4ms time stamp. The triangular mark shows the later point of the
    trajectory, either the end or the last calculated position until that moment. The circular
    mark is the start of the trajectory. We can see that activity moves form CA3 distal to
    proximal in alternating waves of action (hiperpolarization, firing and repolarization).
  }\label{cmevocada}
\end{figure}

\begin{figure}
  \includegraphics[width=0.9\textwidth]{./Figuras/trayevocadasub01.pdf}
  \caption{The evoked trajectories in the r.o.i. We begin to track them
    at the 0.9 ms mark, after letting the saturation of the stimulus recede.
    At 3ms a very clear sink trajectory is moving retrogradly towards CA2,
    sheepherd by two shorter source trayectories. At 5.6 a depolarization
    trajectory starts, following almost the same path and presenting also its
    two companions, ending around the 7.7ms timestamp. No significant
  trajectories are observed after the 10ms mark. }\label{trayevo}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=0.9\textwidth]{./Figuras/LFPvsCSD_4AP.pdf}
  \caption{Comparison of LFP and CSD representations
    at the moments of the epileptic burst. In the first row
    we present the data as measured, no denoising or filtering has
    been applied. The units of measurement are in $\muV{}$. In
    the middle row a direct application of a difference operator
    gives us the CSD. This sort of operators are very sensitive
    to noise, in this case, spatial hard edges, which become
    greatly exaggerated. In the third row we present the same
    data after a Gaussian spatial filter ( Gaussian Blur) has
    been applied. }\label{lfpycsd4ap}
\end{figure}
